---
title: "data_translation_challenge_lpowell"
author: "Lorenzo Powell"
date: "2022-08-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
  Since the onset of the COVID-19 pandemic early 2020, the United States has continued to undergo a labor shortage with rising unemployment rates across all industries. Amidst national lockdowns and a slowing global economy leading to reduced consumer spending, it is expected that the retail sales industry has seen the greatest hardship. The present analysis uses IPUMS Current Population Survey data to investigate the impact of the pandemic on unemployment in retail, insights from the data may help us navigate the upcoming years.  

  Load the necessary packages: 
```{r packages, include = FALSE, echo = FALSE, warning=FALSE, error=FALSE}
library(tidyverse)
library(fixest)
library(vtable)
library(Ecdat)
library(wooldridge)
library(ggplot2)
library(ggstance)
library(NHANES)
library(dagitty)
library(dplyr)
library(purrr)
library(readr)
library(stringr)
library(lubridate)
library(ipumsr)
library(rdrobust)
library(rdd)

```

  Load the data files from github repository and set parameters for mutated columns. Most transformations were made with the intention of better interpreting the column data and to create dummy variables relating to the unemployment & other labor-force status, retail industry, and pre/post COVID-19 pandemic presence. Data was then grouped by year and employment status, not the created dummy-variable.
  For the purposes of this research I've elected to restrict observations to 17 to 64 years old, this should encompass most working age adults; the upper bound for age being 64 was selected as Medicare benefits activate at age 65. IPUMS CPS's employment status responses offers participant feedback for their reasons of not participating in the labor force; while some response options are adequate justification (ie. retirement, inability, school)  others are more vague and can be interpreted as voluntary flight from the work-force. One of the added dummy variables captures responses falling under the 'Not In Labor Force' so that these participants are excluded; because the set cutoff age is 64, respondents stating retirement should be accounted for. 
  The biggest threats to OVB will be the multifaceted impact of both education and healthcare impact. Full/Part time, student expense, and academic rigor will have the greatest internal influence on education whereas contributors to the impact of healthcare are more complex and would require patient health demographics to which we don't have access. Considering these aforementioned shortcomings, valuable inferences can still be made from the following regression models.  
  
```{r Load Data, echo = TRUE, warning=FALSE, error=FALSE, collapse=TRUE}

unemp_fix <- c(20,21,22)
nilf <- c(30:36)

year_precov <- c(2017, 2018, 2019)
retail_ind_precov <- c(1190, 4670:5591)

year_cov <- c(2020, 2021, 2022)
retail_ind_cov <- c(1190, 4670:5790)

high_inc <- (96601:9999999)
med_inc <- (30121:96600)
low_inc <- (1:30120) 
no_inc <- (0)

#NEED TO COMPRESS FILE TO LOAD TO GITHIB

ipums_cps_git <- read_csv('../su_omsba-data_translation_challenge-lpowell/DATA/cps_00004.dat') %>%
rename(HHNUM = SERIAL) %>%
  rename(HHWGHT = HWTFINL) %>%
  rename(NUMEMPSLY = NUMEMPS) %>%
  rename(LWFP = WNLWNILF) %>%
  rename(TPINC = INCTOT) %>%
  rename(TFAMINC = FTOTVAL) %>%
  rename(COMMINS = MRKCOVLY) %>%
  rename(RECSCHYR = EDYRATT) %>%
  rename(REDEGREE = EDYRDEG) %>%
  rename(DIASSISTSS = DIASSISTSSDI) %>%
  rename(DIASSISTO = DIASSISTOTHDI) %>%
  rename(WRKLESSASSIST = DIASSISTWORK) %>%
  mutate(UNEMP = EMPSTAT %in% unemp_fix) %>%
  mutate(NILF = EMPSTAT %in% nilf) %>%
  mutate(PRECOV = YEAR %in% year_precov) %>%
  mutate(POSTCOV = YEAR %in% year_cov) %>%
  filter(AGE > 16, AGE < 65) %>%
  mutate(RETAIL_IND = case_when(
    YEAR %in% year_precov &  as.numeric(IND) %in% retail_ind_precov ~ TRUE,
    YEAR %in% year_cov &  as.numeric(IND) %in% retail_ind_cov ~ TRUE)) %>%
  mutate(UNEMP = EMPSTAT %in% unemp_fix) %>%
  mutate(NILF = EMPSTAT %in% nilf) %>%
  mutate(PRECOV = YEAR %in% year_precov) %>%
  mutate(POSTCOV = YEAR %in% year_cov)  %>%
  group_by(YEAR, EMPSTAT)
ipums_cps$RETAIL_IND[is.na(ipums_cps$RETAIL_IND)] <- FALSE


#REMOVE LOCAL ****
ipums_cps <- read_csv('C:/Users/loren/OneDrive/Desktop/SeattleU/5300 Applied Econometrics/Data Translation Challenge/data_translation_challenge_lpowell/DATA/cps_00004.csv') %>%
  rename(HHNUM = SERIAL) %>%
  rename(HHWGHT = HWTFINL) %>%
  rename(NUMEMPSLY = NUMEMPS) %>%
  rename(LWFP = WNLWNILF) %>%
  rename(TPINC = INCTOT) %>%
  rename(TFAMINC = FTOTVAL) %>%
  rename(COMMINS = MRKCOVLY) %>%
  rename(RECSCHYR = EDYRATT) %>%
  rename(REDEGREE = EDYRDEG) %>%
  rename(DIASSISTSS = DIASSISTSSDI) %>%
  rename(DIASSISTO = DIASSISTOTHDI) %>%
  rename(WRKLESSASSIST = DIASSISTWORK) %>%
  mutate(UNEMP = EMPSTAT %in% unemp_fix) %>%
  mutate(NILF = EMPSTAT %in% nilf) %>%
  mutate(PRECOV = YEAR %in% year_precov) %>%
  mutate(POSTCOV = YEAR %in% year_cov) %>%
  filter(AGE > 16, AGE < 65) %>%
  mutate(RETAIL_IND = case_when(
    YEAR %in% year_precov &  as.numeric(IND) %in% retail_ind_precov ~ TRUE,
    YEAR %in% year_cov &  as.numeric(IND) %in% retail_ind_cov ~ TRUE)) %>%
  mutate(UNEMP = EMPSTAT %in% unemp_fix) %>%
  mutate(NILF = EMPSTAT %in% nilf) %>%
  mutate(PRECOV = YEAR %in% year_precov) %>%
  mutate(POSTCOV = YEAR %in% year_cov)  %>%
  group_by(YEAR, EMPSTAT)
ipums_cps$RETAIL_IND[is.na(ipums_cps$RETAIL_IND)] <- FALSE

#vtable(ipums_cps)


```

  The first 2 regression models included below regress the presence of the pandemic on unemployment, and then controls for age in the latter. Because the medicare eligibility age 65 was chosen as a cutoff, it was best to not include the retirement status of the Not In Labor Force category, doing so allows age cohorts to be compared pre & post pandemic labor trends without superfluous respondents.  The observations reported are significant at the 1% level and indicate a positive association between the variables for pandemic presence and employment in the retail industry. The first model indicates that the presence of the pandemic increases unemployment by 0.0047 units. When adding age as another variable the pandemic's presence is associated with a -0.0007 unit change in age of the unemployed employee. The regressed effects are not as impressive as expected but in my exploratory analysis I found regression continuity models to fit well with the design of my data transformations. 

```{r Industry & Unemployment, echo = TRUE, warning=FALSE, error=FALSE}
ind_reg1 <- feols(UNEMP ~ POSTCOV*RETAIL_IND, data = ipums_cps)
ind_reg2 <- feols(UNEMP ~ POSTCOV*RETAIL_IND + AGE, data = ipums_cps)
etable(ind_reg1, ind_reg2)

```

  Despite the reduction in statistical power, the more optimal models use regression discontinuity with a cutoff of the pandemic start year, 2020. These models examine average retail employee hours and overall employment have fared before and after 2020. Under both dependent metrics, we observe a positive correlation at the 1% significance level between onset of the pandemic and the retail industry. After 2020, a 1 unit change in year is associated with a 0.0204 unit increase in unemployment and a -218.9 unit change in average weekly hours worked.  
  The third regression adds controls for both age and FTE/PTE employment status to the prior model; with these controls in place, retail industry employees are likely to work 95.58 units less after the onset of the pandemic, and a 1 unit increase in age is associated with a -2.672 unit change in weekly hours worked. 
  These models support the assumptions that retailers are disproportionately impacted by staffing following the pandemic onset, and that younger employees are more likely to reduce or remove their presence in the work force.

```{r Industry % Unemployment/Hours, echo = TRUE, warning=FALSE, error=FALSE}
rdd_retail1 <- feols(UNEMP ~ RETAIL_IND + I(YEAR - 2020), data = ipums_cps)
rdd_retail_hrs <- feols(UHRSWORK1 ~ RETAIL_IND + I(YEAR - 2020), data = ipums_cps)
rdd_retail_hrs2 <- feols(UHRSWORK1 ~ RETAIL_IND + AGE + FULLPART + I(YEAR - 2020), data = ipums_cps)
etable(rdd_retail1, rdd_retail_hrs,rdd_retail_hrs2)

```

  For the participants who have degrees, there is a 0.0001 unit increase in unemployment in the same year of earning that degree when controlling for age. Age is negatively associated with unemployment where there is a -0.0007 unit change in age while controlling for recency of degree. Unfortunately after excluding NA values only 279,139 remain, so the findings should only be used as ancillary details to bolster other regression observations. 

```{r Education & Unemployment, echo = TRUE, warning=FALSE, error=FALSE}
unemp_ed <-feols(UNEMP ~ REDEGREE + AGE, data = ipums_cps)
etable(unemp_ed)

```
  The last cluster investigates retail employee absenteeism and average weekly hours worked in the setting of the pandemic. The first model indicates that without any additional controls or consideration of the pandemic a 1 unit change in paid absence is associated with a 1.37 unit increase in absenteeism. This can be understood as an incentive for staff to continue the unreliable behavior.
    The second model regresses absenteeism on payment for absenteeism, and adds controls for an employees FTE/PTE status and age. As expected FTE/PTE status had the greatest impact on days absent from work. While controlling for FTE/PTE status, and age, paid absence is associated with a -48.57 unit change in average weekly hours worked at the 1% significance level. This is interpreted as paid absenteeism being strongly correlated with reduced average weekly hours worked. 
    The last model regresses the average weekly hours worked on pandemic presence in the retail industry on. While the joint effect was not statistically significant the other findings in this model are of interest as they contrast expectations in part; but used with data from other models may provide some insight on the overall labor market landscape. The model indicates that, while controlling for working in the retail industry, the presence of the pandemic is associated with a 14.09 unit increase in average hours worked; controlling for the pandemic, retail industry employees see a -219.9 unit decrease in working hours. 

```{r Paid Absenteeism, echo = TRUE, warning=FALSE, error=FALSE}
absent_pay <- feols(ABSENT ~ PAYIFABS, data = ipums_cps)
absent_pay2 <- feols(UHRSWORK1 ~ PAYIFABS + FULLPART + AGE, data = ipums_cps)
retail_hrs <- feols(UHRSWORK1 ~ POSTCOV*RETAIL_IND, data = ipums_cps)
etable(absent_pay, absent_pay2, retail_hrs)

```
